---
title: "Set the data directory"
vignette: >
  %\VignetteIndexEntry{Set the data directory}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute: 
  eval: false
---


## Set the data directory {#set-data-folder}

Choose where `{spanishoddata}` should download (and convert) the data by setting the `SPANISH_OD_DATA_DIR` environment variable with the following command:

```{r}
Sys.setenv(SPANISH_OD_DATA_DIR = "~/spanish_od_data")
```

The package will create this directory if it does not exist on the first run of any function that downloads the data.

<details><summary>Setting data directory for advanced users</summary>

To permanently set the directory for all projects, you can specify the data directory globally by setting the `SPANISH_OD_DATA_DIR` environment variable, e.g. with the following command:

```{r}
#| eval: false
usethis::edit_r_environ()
# Then set the data directory globally, by typing this line in the file:
```

```
SPANISH_OD_DATA_DIR = "~/spanish_od_data"
```

You can also set the data directory locally, just for the current project. Set the 'envar' in the working directory by editing `.Renviron` file in the root of the project:

```{r}
#| eval: false
file.edit(".Renviron")
```


</details>
